EBASE manuscript

### 2023-02-26 manu-draft.qmd
update apasumdat figure and text for mean and sd prior comps

\newcommand{\auni}{\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)}
\newcommand{\runi}{\text{mmol}/\text{m}^3/\text{d}}
\newcommand{\buni}{\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)}

## Abstract {.unnumbered}

@@

* Overview of metabolism - importance
* Existing methods: Odum/WtRegDO, BASE, references in @Grace15, applications to lakes and streams
* Applications to estuaries - challenges
* Goals and objectives

@@

where the terms in the square brackets are production (P), respiration (R), and gas exchange (D), respectively, modeled as a function of the change in dissolved oxygen ($\partial{C_d}$) per unit time ($\partial{t}$) of the input data.  Production is modeled linearly as a function of surface irradiance ($PAR$, Watts m$^{-2}$) and $a$ (light efficiency, $\auni$).  Respiration is modeled with the single parameter $r$ ($\runi$).  Gas exchange is modeled using the formulation in @Wanninkhof14 as a wind-based parameterization for the gas transfer velocity [@Ho16] multiplied by the oxygen concentration gradient as the difference between the concentration measured in the surface water layer, $C_d$ (mmol m$^{-3}$), and the saturation (or equilibrium) concentration, $C_{Sat}$ (mmol m$^{-3}$), calculated from salinity (psu) and temperature ($^\circ$C) [@Millero81; @Garcia92]. The gas transfer velocity is modeled using wind speed squared at 10 meters above the water ($U_{10}^2$, m$^2$ s$^{-2}$), water column depth ($H$ in meters, as a fixed parameter or time series from a pressure sensor), and the Schmidt number ($S_c$) defined as the ratio of the kinematic viscosity of water to the molecular diffusivity of the gas for oxygen [from water temperature and salinity using the polynomial fit in @Wanninkhof14].  The $b$ parameter is estimated from the data and has units of $\buni$.  As such, the three parameters estimated by EBASE are $a$, $r$, and $b$ using the required input data that includes time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth (`r run_reference(id = "ebaseiotab")`).

@@

A common theory behind most gas exchange parameterizations for metabolism estimates is that the transfer of gases of relatively low solubility, like oxygen, is proportional to the concentration gradient in a thin film of water at the air–water interface. It is also assumed that the water at the top of the film is saturated with respect to the atmosphere, so that the concentration gradient across the thin film can be approximated as the difference between oxygen concentration measured in the surface water layer, $C_d$, and the saturation (or equilibrium) concentration $C_{Sat}$, which is usually calculated as a function of water temperature and salinity (the negative sign is needed because positive flux is set to be from water to air):

@@

The gas transfer velocity, $k_w$, has dimensions of length over time and increases as the water column in the vicinity of the air–water interface becomes more turbulent.  As such, $k_w$  is usually modeled as a function of measurable variables that can potentially predict turbulent mixing of the surface water, e.g., wind speed, current velocity, and water depth [@Ho16]. $k_w$ also increases with temperature because the film of water at the air-sea interface becomes more thin as the random motion of gas and water molecules increases at higher temperatures and transfer of molecules across the film increases. This dependence of $k_w$ on temperature is usually modeled with the Schmidt number as in @eq-ebase, which combines information on the viscosity of the fluid and the diffusivity of the gas.

@@

where $\bar{T}$ is the average water temperature over each 24-hour period in the diel oxygen time series and $K$ is the reaeration coefficient estimated from the data.  @eq-gasexbase indicates that $k_w$ increases with temperature and depth. The temperature dependence is consistent with the known decrease of $k_w$ with the Schmidt number, but the depth dependence is difficult to justify.  An expectation is that $k_w$ would increase as water depth decreases for a given current speed because turbulence generated at the sediment–water interface by currents would more easily reach the surface [@Ho16].  Based on these differences, the wind-based parameterization for gas exchange was used for EBASE  [@Wanninkhof14].  Additionally, $C_{sat}$ was calculated using only temperature and salinity, removing the need for barometric pressure used in BASE.

The additional modifications to @eq-base for EBASE were removing the $p$ and $\Theta$ parameters from the production (P) and respiration (R) components, respectively.  The BASE method provides options to estimate both $\Theta$ and $p$ from the data or to set as fixed values defined in the model file ($\Theta = 1.07177$, $p=1$). For EBASE, the temperature dependency of respiration is removed by setting $\Theta^{(T-\bar{T})}$ = 1 and it is assumed that respiration is constant as in other methods [@Odum56;@Murrel18].  The $p$ parameter, which is dimensionless, is also set to 1 as in @Holtgrieve10, such that a linear relationship of production to $PAR$ was used.  This removes the exponential dependence of production with increasing $PAR$. Removing these components also improved model stability by creating more reasonable estimates for the other parameters.

### EBASE model estimation

The remaining three parameters $a$ (light efficiency, $\auni$), $r$ (respiration, $\runi$), and $b$ ($\buni$) are estimated by likelihood given the observed data and prior distributions for the parameters.  The JAGS software [as for BASE, @Plummer03] is used with the EBASE R package [described below, @Beck22;@RCT22] to estimate the unknown parameters using Markov Chain Monte Carlo simulations in a Bayesian framework. At each time step, the change in oxygen concentration between time steps is calculated from the equation using model inputs and parameter guesses based on the prior distributions, and then a finite difference approximation is used to estimate modeled oxygen concentration. Estimations proceed using a forward in time integration where the dissolved oxygen at each time step is informed by parameter estimates from the preceding time steps.  The estimated dissolved oxygen concentration is also returned, which can be compared to the observed values as one measure of model performance.  All parameter inputs and outputs for EBASE are shown in `r run_reference(id = "ebaseiotab")`.  Areal rates for $P$, $R$, and $D$ (mmol m$^-2$ d$^{-1}$) are returned for comparability of estimates across locations where water column depth may vary.

The metabolism estimates and their parameters returned by the Bayesian routine implemented in JAGS are affected by the prior distributions assigned to each.  As for the BASE model, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can be changed based on previous knowledge of parameters specific to an ecosystem or as informed by other metabolic modeling approaches.  The default priors were based on those that returned reasonable parameter estimates using a known time series created with forward-feeding oxygen mass balance inputs to @eq-ebase.  [Fwoxy R package, @Arriola22]. All of the physical parameters were held constant, i.e., wind speed, temperature, and salinity.  The other required inputs were the $a$, $r$, and $b$ parameters, chosen as 0.2 $\auni$, 20 $\runi$, and 0.251 $\buni$, respectively.  An initial oxygen concentration was provided as 250 mmol/m$^3$ and the time series was created by forward addition of the inputs. As such, the dissolved oxygen output as change per unit time was modelled by EBASE using the constant the same physical parameters as input. Reasonable prior distributions by EBASE were identified (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]) based on the ability to return the known parameters:

@@

Each follows a normal distribution with means and standard deviation defined in the parentheses (@fig-priorplot). The $T$ notation is JAGS model syntax indicating the distributions are truncated to a minimum (left) and maximum (right) value defined in the parentheses.  All prior distributions were constrained to positive values based on known physical constraints and requirements of the model formula in @eq-ebase.  Additionally, the prior distribution for $b$ was constrained to an upper limit of 0.504 (~twice the default mean).  @Wanninkhof14 states that the $b$ parameter typically does not vary by more than 20% and the standard deviation of the default prior is sufficiently high to expand beyond this range. Initial development of EBASE also showed that using an undefined upper limit of $b$ led to unstable and unreasonable parameter estimates.

For ease of application, an R package was developed to implement the EBASE methods described above [@Beck22]. A detailed web page (<https://fawda123.github.io/EBASE/>) describes the required data inputs for estimating the metabolic parameters using the core R functions that pass inputs to the JAGS software.  A JAGS model file is included with the package that implements @eq-ebase, with options to supply a custom model file as needed.  Users can also specify alternative prior distributions from those in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] for each parameter. Functions are available for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D, as areal rates with units $\runi$), and a time series plot of the credible intervals for the $a$, $r$, and $b$ parameters. An additional option also allows the user to specify the model optimization period, where the time period can vary from a minimum of one day to the maximum of the entire length of the time series. This differs from the BASE approach where the optimization period is set at one day. Sensitivity of the EBASE method to different optimization periods is described in @sec-cmpmod. All results provided herein were generated using the EBASE R package.

@@

A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $r$, and $b$ (@eq-ebase) was conducted using a synthetic time series with actual physical data as inputs. These assessments evaluated if 1) EBASE estimates similar values for the known parameters, and 2) how the estimated values change with different model configurations. The synthetic time series used inputs from an approximate one year of continuous data from Apalachicola Bay, Florida.  The input data were 2012 observations from February 22nd to December 19th at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, -84.8802$^{\circ}$ W) and included 15 minute observations of dissolved oxygen (mg/L), water temperature (C), and salinity (psu).  Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, -84.8834$^{\circ}$ W, ~10km north of Cat Point) were combined with the water quality data at the same time step and included air temperature (C), wind speed (m/s), and PAR (mmol/m$^2$ as total per 15 minute observation). An oxygen time series was generated using the inputs from Apalachicola Bay ([@fig-synapa]a), such that the synthetic dissolved oxygen was created as close as possible to the observed given the inputs.  Known values for $a$ and $r$ parameters were returned for each time step of the synthetic data, as well as all metabolic estimates for production, respiration, and gas exchange ([@fig-synapa]b and c).  The only parameter that could not be created as a continuous time series was $b$, which was set to 0.251 $\buni$ [@Wanninkhof14].

@@

The standard prior distributions for the EBASE parameters follow normal distributions with defined means and standard deviations (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]).  The sensitivity analysis evaluated changes in the distributions by varying the standard deviations from small to large to test the effect of more or less constrained ranges, respectively, on the priors. For each parameter, low, medium, and high values for the standard deviation were assessed as 10%, 100%, and 1000% of the default values.  Each unique combination of low, medium, and high values for the standard deviation of each parameter were evaluated, creating 3$^3$ = 27 different combinations of prior distributions that were used with the synthetic time series in EBASE.

The optimization period used for a time series determines how much data are used to estimate the three EBASE parameters ($a$, $r$, and $b$).  Unlike BASE, the EBASE software allows flexibility in the period of time used for optimization, with a minimum period of time being one day and a maximum period being the length of the time series.  For a chosen optimization period, single estimates for $a$, $r$ and $b$ are returned by the model.  A short period of optimization (e.g., one day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE.  As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation.  An optimization period of one day was compared to results using an optimization period of seven days for the synthetic one-year time series. Each optimization period was evaluated for every unique combination of prior distributions, totaling 54 (27 times the two optimization periods) unique analyses for the assessment of changing standard deviations of all parameters.

Changing the variance of each of the prior distributions affected the ability of EBASE to return the known parameters from the synthetic time series. @fig-apasumdat shows a summary of the comparisons of the synthetic data with EBASE for each unique combination of prior distributions and two optimization periods of one and seven days.  The results are compared using R-squared ($R^2$) values from the linear fit between the synthetic data and results from EBASE.  Darker colors represent lower $R^2$ values and can be used to assess how changing the variance of the priors affects the ability of EBASE to estimate the true parameter.

@@

The results suggest that the default prior distributions used in EBASE may be appropriate in some cases, but increasing the prior variance for $a$ and $r$ and decreasing the variance for $b$ will likely produce more optimal solutions. The actual model results for the best performing combination of prior distributions for the one day and seven day optimization are shown in @fig-optex (all results are averaged at the time-step of the optimization period for comparability).  The best performing models were those with the highest mean $R^2$ values across all parameter estimates in @fig-apasumdat.  For both models, the metabolic parameters were estimated with high precision, and in some cases with nearly perfect precision (i.e., dissolved oxygen and gas exchange, D).  Estimates for the fixed $b$ parameter were most variable, but not biased as the mean for both models was close to 0.251 $\buni$.

@@

Metabolic estimates from EBASE were further compared to those from existing methods to benchmark and evaluate potential differences from alternative model formulations used by each approach. The observed one year continuous time series described above for water quality and weather data from Apalachicola Bay was used to compare metabolic estimates from EBASE, BASE, and the Odum open-water method [@Odum56]. The Odum method used an approach described in @Murrell18 created for estuarine application that uses a gas-exchange approach described in @Thebault08.  For better comparison with the EBASE methods, the gas exchange parmaterization in the Odum method was modified to use the @Wanninkhof14 approach [as available in the WtRegDO R package, @Beck21]. The input data were provided to each method after converting the units as required (e.g., BASE requires PAR as $\mu$mol/m$^{2}$/s).  Water column depth at Cat Point is approximately 1.8 m and all daily metabolic estimates returned by each method were converted to volumetric rates by dividing by the depth (i.e., $\runi$).  Each comparison also evaluated metabolic estimates using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; available from @Beck21].  The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection.  All metabolic estimates were obtained using the default settings for each of the three methods, with an exception that the prior distribution for the reaeration coefficient ($K$) in BASE was informed by results from the Odum output.  Initial assessments indicated that unstable and inaccurate results for the time series are obtained from BASE using an uninformed prior distribution for $K$.  All other prior distributions for BASE and EBASE were set as the software default.  A one day model optimization period was used for EBASE for comparability with BASE.

@@

* Describe how EBASE deals with gaps

@@

![Comparison of EBASE results to a synthetic time series created with known metabolic parameters. Results are compared for unique combination of prior distributions, where the standard deviations are evaluated from low (L), medium (M), to high (H) for a one day and seven day model optimization period. The metabolic results for dissolved oxygen (DO), production (P), respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on the $R^2$ value of the linear fit between the EBASE results and synthetic values.  The $b$ parameter is a fixed constant and was not evaluated.](figs/apasumdat.png){#fig-apasumdat}

@@

![Comparison of EBASE results for the best performing model to a synthetic time series created with known metabolic parameters. The best performing model for the one day (a, ndays = 1) and seven day (b, ndays = 7) optimization period is based on the EBASE results with prior distributions that produced the highest mean $R^2$ values across all parameter estimates (@fig-apasumdat).  The standard deviations of the prior distributions of each parameter for the best performing model are shown at the top.  The results are averaged at the time step of the model optimization period.  The $b$ parameter is fixed at 0.251 $\buni$.](figs/optex.png){#fig-optex fig-align="center"}

>>>>

```{=tex}
\newcommand{\auni}{\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)}
\newcommand{\runi}{\text{mmol}/\text{m}^3/\text{d}}
\newcommand{\buni}{\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)}
```
## Abstract {.unnumbered}

@@

-   Overview of metabolism - importance
-   Existing methods: Odum/WtRegDO, BASE, references in @Grace15, applications to lakes and streams
-   Applications to estuaries - challenges
-   Goals and objectives

@@

where the terms in the square brackets are production (P), respiration (R), and gas exchange (D), respectively, modeled as a function of the change in dissolved oxygen ($\partial{C_d}$) per unit time ($\partial{t}$) of the input data. Production is modeled linearly as a function of surface irradiance ($PAR$, Watts m$^{-2}$) and $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$). Respiration is modeled with the single parameter $r$ ($\text{mmol}/\text{m}^3/\text{d}$). Gas exchange is modeled using the formulation in @Wanninkhof14 as a wind-based parameterization for the gas transfer velocity [@Ho16] multiplied by the oxygen concentration gradient as the difference between the concentration measured in the surface water layer, $C_d$ (mmol m$^{-3}$), and the saturation (or equilibrium) concentration, $C_{Sat}$ (mmol m$^{-3}$), calculated from salinity (psu) and temperature ($^\circ$C) [@Millero81; @Garcia92]. The gas transfer velocity is modeled using wind speed squared at 10 meters above the water ($U_{10}^2$, m$^2$ s$^{-2}$), water column depth ($H$ in meters, as a fixed parameter or time series from a pressure sensor), and the Schmidt number ($S_c$) defined as the ratio of the kinematic viscosity of water to the molecular diffusivity of the gas for oxygen [from water temperature and salinity using the polynomial fit in @Wanninkhof14]. The $b$ parameter is estimated from the data and has units of $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$. As such, the three parameters estimated by EBASE are $a$, $r$, and $b$ using the required input data that includes time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth (`r run_reference(id = "ebaseiotab")`).

@@

A common theory behind most gas exchange parameterizations for metabolism estimates is that the transfer of gases of relatively low solubility, like oxygen, is proportional to the concentration gradient in a thin film of water at the air--water interface. It is also assumed that the water at the top of the film is saturated with respect to the atmosphere, so that the concentration gradient across the thin film can be approximated as the difference between oxygen concentration measured in the surface water layer, $C_d$, and the saturation (or equilibrium) concentration $C_{Sat}$, which is usually calculated as a function of water temperature and salinity (the negative sign is needed because positive flux is set to be from water to air):

@@

The gas transfer velocity, $k_w$, has dimensions of length over time and increases as the water column in the vicinity of the air--water interface becomes more turbulent. As such, $k_w$ is usually modeled as a function of measurable variables that can potentially predict turbulent mixing of the surface water, e.g., wind speed, current velocity, and water depth [@Ho16]. $k_w$ also increases with temperature because the film of water at the air-sea interface becomes more thin as the random motion of gas and water molecules increases at higher temperatures and transfer of molecules across the film increases. This dependence of $k_w$ on temperature is usually modeled with the Schmidt number as in @eq-ebase, which combines information on the viscosity of the fluid and the diffusivity of the gas.

@@

where $\bar{T}$ is the average water temperature over each 24-hour period in the diel oxygen time series and $K$ is the reaeration coefficient estimated from the data. @eq-gasexbase indicates that $k_w$ increases with temperature and depth. The temperature dependence is consistent with the known decrease of $k_w$ with the Schmidt number, but the depth dependence is difficult to justify. An expectation is that $k_w$ would increase as water depth decreases for a given current speed because turbulence generated at the sediment--water interface by currents would more easily reach the surface [@Ho16]. Based on these differences, the wind-based parameterization for gas exchange was used for EBASE [@Wanninkhof14]. Additionally, $C_{sat}$ was calculated using only temperature and salinity, removing the need for barometric pressure used in BASE.

The additional modifications to @eq-base for EBASE were removing the $p$ and $\Theta$ parameters from the production (P) and respiration (R) components, respectively. The BASE method provides options to estimate both $\Theta$ and $p$ from the data or to set as fixed values defined in the model file ($\Theta = 1.07177$, $p=1$). For EBASE, the temperature dependency of respiration is removed by setting $\Theta^{(T-\bar{T})}$ = 1 and it is assumed that respiration is constant as in other methods [@Odum56; @Murrel18]. The $p$ parameter, which is dimensionless, is also set to 1 as in @Holtgrieve10, such that a linear relationship of production to $PAR$ was used. This removes the exponential dependence of production with increasing $PAR$. Removing these components also improved model stability by creating more reasonable estimates for the other parameters.

### EBASE model estimation {#sec-ebasemod}

The remaining three parameters $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$), $r$ (respiration, $\text{mmol}/\text{m}^3/\text{d}$), and $b$ ($\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$) are estimated by likelihood given the observed data and prior distributions for the parameters. The JAGS software [as for BASE, @Plummer03] is used with the EBASE R package [described below, @Beck22; @RCT22] to estimate the unknown parameters using Markov Chain Monte Carlo simulations in a Bayesian framework. At each time step, the change in oxygen concentration between time steps is calculated from the equation using model inputs and parameter guesses based on the prior distributions, and then a finite difference approximation is used to estimate modeled oxygen concentration. Estimations proceed using a forward in time integration where the dissolved oxygen at each time step is informed by parameter estimates from the preceding time steps. The estimated dissolved oxygen concentration is also returned, which can be compared to the observed values as one measure of model performance. All parameter inputs and outputs for EBASE are shown in `r run_reference(id = "ebaseiotab")`. Areal rates for $P$, $R$, and $D$ (mmol m$^-2$ d$^{-1}$) are returned for comparability of estimates across locations where water column depth may vary.

The metabolism estimates and their parameters returned by the Bayesian routine implemented in JAGS are affected by the prior distributions assigned to each. As for the BASE model, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can be changed based on previous knowledge of parameters specific to an ecosystem or as informed by other metabolic modeling approaches. The default priors were based on those that returned reasonable parameter estimates using a known time series created with forward-feeding oxygen mass balance inputs to @eq-ebase. [Fwoxy R package, @Arriola22]. All of the physical parameters were held constant, i.e., wind speed, temperature, and salinity. The other required inputs were the $a$, $r$, and $b$ parameters, chosen as 0.2 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$, 20 $\text{mmol}/\text{m}^3/\text{d}$, and 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$, respectively. An initial oxygen concentration was provided as 250 mmol/m$^3$ and the time series was created by forward addition of the inputs. As such, the dissolved oxygen output as change per unit time was modelled by EBASE using the constant the same physical parameters as input. Reasonable prior distributions by EBASE were identified (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]) based on the ability to return the known parameters:

@@

Each follows a normal distribution with means and standard deviation defined in the parentheses (@fig-priorplot). The $T$ notation is JAGS model syntax indicating the distributions are truncated to a minimum (left) and maximum (right) value defined in the parentheses. All prior distributions were constrained to positive values based on known physical constraints and requirements of the model formula in @eq-ebase. Additionally, the prior distribution for $b$ was constrained to an upper limit of 0.504 (\~twice the default mean). @Wanninkhof14 states that the $b$ parameter typically does not vary by more than 20% and the standard deviation of the default prior is sufficiently high to expand beyond this range. Initial development of EBASE also showed that using an undefined upper limit of $b$ led to unstable and unreasonable parameter estimates.

For ease of application, an R package was developed to implement the EBASE methods described above [@Beck22]. A detailed web page (<https://fawda123.github.io/EBASE/>) describes the required data inputs for estimating the metabolic parameters using the core R functions that pass inputs to the JAGS software. A JAGS model file is included with the package that implements @eq-ebase, with options to supply a custom model file as needed. Users can also specify alternative prior distributions from those in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] for each parameter. Functions are available for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D, as areal rates with units $\text{mmol}/\text{m}^3/\text{d}$), and a time series plot of the credible intervals for the $a$, $r$, and $b$ parameters. An additional option also allows the user to specify the model optimization period, where the time period can vary from a minimum of one day to the maximum of the entire length of the time series. This differs from the BASE approach where the optimization period is set at one day. Sensitivity of the EBASE method to different optimization periods is described in @sec-cmpmod. All results provided herein were generated using the EBASE R package.

@@

A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $r$, and $b$ (@eq-ebase) was conducted using a synthetic time series with actual data as inputs. These assessments evaluated if 1) EBASE estimates similar values for the known parameters, and 2) how the estimated values change with different model configurations. The synthetic time series used inputs from an approximate one year of continuous data from Apalachicola Bay, Florida. The input data were 2012 observations from February 22nd to December 19th at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, -84.8802$^{\circ}$ W) and included 15 minute observations of dissolved oxygen (mg/L), water temperature (C), and salinity (psu). Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, -84.8834$^{\circ}$ W, \~10km north of Cat Point) were combined with the water quality data at the same time step and included air temperature (C), wind speed (m/s), and PAR (mmol/m$^2$ as total per 15 minute observation). An oxygen time series was generated using the inputs from Apalachicola Bay ([@fig-synapa]a), such that the synthetic dissolved oxygen was created as close as possible to the observed given the inputs. Known values for $a$ and $r$ parameters were returned for each time step of the synthetic data, as well as all metabolic estimates for production, respiration, and gas exchange ([@fig-synapa]b and c). The only parameter that could not be created as a continuous time series was $b$, which was set to 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [@Wanninkhof14].

@@

The standard prior distributions for the EBASE parameters follow normal distributions with defined means and standard deviations (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]). The sensitivity analysis evaluated changes in the distributions by varying the means and standard deviations from small to large to test the effect of changes in the central tendencies and more or less constrained ranges, respectively, on the priors. For each parameter, low and high values for the means and standard deviation were assessed as changes approximately equal to two orders of magnitude. For example the default value for the mean of the $a$ prior distribution is 0.2 and low and high values were assessed as 0 and 2, respectively. The only exception was the high value that was tested for the mean of the $b$, which was set at 0.504 or twice the default value (justification noted in @sec-ebasemod). Each unique combination of low and high values for the means and standard deviations of each parameter were evaluated, creating 2$^6$ = 64 different combinations of prior distributions that were used with the synthetic time series in EBASE.

The optimization period used for a time series determines how much data are used to estimate the three EBASE parameters ($a$, $r$, and $b$). Unlike BASE, the EBASE software allows flexibility in the period of time used for optimization, with a minimum period of time being one day and a maximum period being the length of the time series. For a chosen optimization period, single estimates for $a$, $r$ and $b$ are returned by the model. A short period of optimization (e.g., one day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE. As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation. An optimization period of one day was compared to results using an optimization period of seven days for the synthetic one-year time series. Each optimization period was evaluated for every unique combination of prior distributions, totaling 128 (64 times the two optimization periods) unique analyses for the assessment of changing priors of all parameters.

Changing the prior distributions affected the ability of EBASE to return the known parameters from the synthetic time series, with results varying by optimization period. @fig-apasumdat shows a summary of the comparisons of the synthetic data with EBASE for each unique combination of prior distributions and two optimization periods of one and seven days. The results are compared using R-squared ($R^2$) values from the linear fit between the synthetic data and results from EBASE. Darker colors represent lower $R^2$ values and can be used to assess how changing the variance of the priors affects the ability of EBASE to estimate the true parameter.

@@

The results suggest that the default prior distributions used in EBASE may be appropriate in some cases, but increasing the prior variance for $a$ and $r$ and decreasing the variance for $b$ will likely produce more optimal solutions. The actual model results for the best performing combination of prior distributions for the one day and seven day optimization are shown in @fig-optex (all results are averaged at the time-step of the optimization period for comparability). The best performing models were those with the highest mean $R^2$ values across all parameter estimates in @fig-apasumdat. For both models, the metabolic parameters were estimated with high precision, and in some cases with nearly perfect precision (i.e., dissolved oxygen and gas exchange, D). Estimates for the fixed $b$ parameter were most variable, but not biased as the mean for both models was close to 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$.

@@

Metabolic estimates from EBASE were further compared to those from existing methods to benchmark and evaluate potential differences from alternative model formulations used by each approach. The observed one year continuous time series described above for water quality and weather data from Apalachicola Bay was used to compare metabolic estimates from EBASE, BASE, and the Odum open-water method [@Odum56]. The Odum method used an approach described in @Murrell18 created for estuarine application that uses a gas-exchange approach described in @Thebault08. For better comparison with the EBASE methods, the gas exchange parmaterization in the Odum method was modified to use the @Wanninkhof14 approach [as available in the WtRegDO R package, @Beck21]. The input data were provided to each method after converting the units as required (e.g., BASE requires PAR as $\mu$mol/m$^{2}$/s). Water column depth at Cat Point is approximately 1.8 m and all daily metabolic estimates returned by each method were converted to volumetric rates by dividing by the depth (i.e., $\text{mmol}/\text{m}^3/\text{d}$). Each comparison also evaluated metabolic estimates using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; available from @Beck21]. The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection. All metabolic estimates were obtained using the default settings for each of the three methods, with an exception that the prior distribution for the reaeration coefficient ($K$) in BASE was informed by results from the Odum output. Initial assessments indicated that unstable and inaccurate results for the time series are obtained from BASE using an uninformed prior distribution for $K$. All other prior distributions for BASE and EBASE were set as the software default. A one day model optimization period was used for EBASE for comparability with BASE.

@@

-   Describe how EBASE deals with gaps

@@

![Comparison of EBASE results to a synthetic time series created with known metabolic parameters. Results are compared for unique combination of prior distributions, where the means and standard deviations are evaluated from low (L) to high (H) for a one day and seven day model optimization period. The metabolic results for dissolved oxygen (DO), production (P), respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on the $R^2$ value of the linear fit between the EBASE results and synthetic values. The $b$ parameter is a fixed constant and was not evaluated.](figs/apasumdat.png){#fig-apasumdat}

@@

![Comparison of EBASE results for the best performing model to a synthetic time series created with known metabolic parameters. The best performing model for the one day (a, ndays = 1) and seven day (b, ndays = 7) optimization period is based on the EBASE results with prior distributions that produced the highest mean $R^2$ values across all parameter estimates (@fig-apasumdat). The standard deviations of the prior distributions of each parameter for the best performing model are shown at the top. The results are averaged at the time step of the model optimization period. The $b$ parameter is fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$.](figs/optex.png){#fig-optex fig-align="center"}
